Climatic warming strengthens a positive feedback between alpine shrubs and fire

### 2016-03-07 manuscript/manuscript.Rnw
Filled in most key values. Still more to do

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment into arctic and alpine grasslands. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to 36 fold) and that warming doubled tall shrub seedling growth rates and may increase survival. Warming had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increase survival.
\end{abstract}

@@

\subsection{Drivers of shrub seedling establishment.}

We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (two dominant shrubs of open heathland). In 2012-13, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between 13 and 36 times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were 1.44/m$^2$ at burnt sites and 0.04/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: 1.71 and 0.13 seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a 200 m altitudinal range (equivalent to a 1.5\textdegree C change in mean temperature). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a 1\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

Knowing shrub seedlings were more abundant in burnt vegetation, we investigated whether warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} emerging after a wildfire were transplanted into experimentally burnt plots that were either passively warmed by 1\textdegree C using open top chambers (OTCs) or subject to current ambient conditions (controls). After 1812 days (1087 growing season days where OTC treatment was applied) or 5 years growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium}) growing in post-fire bare ground were taller in warmed plots by 9.7, 13.5 and 3.8 cm respectively (Fig. \ref{fig3}). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no significant difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, mean annual predicted height growth rates of \textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium} were 2.4, 1.9 and 1.5 times that observed in control plots, respectively. Projecting these differences equates to an average 6 cm tall seedling (the mean initial size used in this experiment) growing in warmed conditions reaching maximum mature height in 39 fewer years for \textit{Grevillea} and 17 fewer years for both \textit{Prostanthera} and \textit{Phebalium} (Fig. \ref{fig3}). \textit{Asterolasia} was predicted to reach a mature height in approximately 22 years regardless of warming treatment.

We validated the \textit{Grevillea} and \textit{Asterolasia} growth responses observed in experimentally warmed plots by using an elevation gradient in open heathland and maximum height data for 2003 post-fire recruits (i.e. suspected 9 year old seedlings). Here, a 200 m altitudinal range is equivalent to an ambient temperature difference of approximately 1.5\textdegree C\cite{Brown:1989vx}, which is comparable to that observed between experimental warmed and control plots (1\textdegree C). In response to shifts in temperature, maximum heights of Grevillea post-fire recruits were expected to decrease with elevation, whereas \textit{Asterolasia} seedlings were not expected to show this pattern. Our experimental predictions were verified (Supplementary Fig. S4). Mean maximum heights of \textit{Grevillea} seedlings in burnt open heathland were 7.9 cm taller at 1660 m a.s.l compared to seedlings at 1860 m a.s.l. (19.4 cm vs 11.5 cm; a difference comparable to our experimental findings). In contrast, mean maximum heights of \textit{Asterolasia} seedlings did not vary significantly with elevation. Topographic Wetness Index (TWI) and fire severity had no detectable influence on maximum seedling heights in either species.

@@

After five years and across all plots, 34\% (172 out of 512) of all seedlings growing in burnt bare ground patches had died. Most deaths occurred in \textit{Asterolasia} (65) followed by \textit{Phebalium} (63), \textit{Grevillea} (31) and \textit{Prostanthera} (13). \textit{Prostanthera} showed the largest treatment effect (Fig. \ref{fig4}), with annual mortality rates estimated to be near 0 in warmed plots and 5\% in control plots. This significant increase in survival may be a consequence of OTCs reducing the severity of spring frosts by rising minimum ambient and soil temperatures by 0.86\textdegree C and 1.6\textdegree C, respectively (see Supplementary Fig S5-7). Warming also reduced mean seedling mortality in \textit{Grevillea} and \textit{Phebalium} (Fig. \ref{fig4}); however, for both species, the effect was not statistically significant. By contrast, annual mortality rates in the prostrate shrub, \textit{Asterolasia}, were marginally higher in warmed plots, but again this effect was not significant (Fig. \ref{fig4}).

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still reduce shrub recruitment opportunities by infilling post-fire bare ground gaps at a faster rate. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses. However, the rate at which this occurred was not significantly influenced by a 1\textdegree C increase in temperature (Fig. \ref{fig6}).

@@

\begin{figure}
\label{fig1}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig1.pdf}
\caption{\textbf{Conceptual diagram illustrating positive warming-shrub-fire feedback loop.} Top panel highlights the known positive feedback loop between fire, climatic warming and landscape flammability (solid blue lines). It also highlights several interactions (dashed \& black) which currently have a paucity of information but which may exacerbate or diminish this feedback loop. Bottom panel either assigns the direction or removes these interactions based on new evidence obtained from experimental and field surveys conducted in this study. It also includes an example of a potential, but unlikely, effect (red dotted line) that could break this feedback.}
\end{minipage}
\end{figure}

\begin{figure}
\label{fig2}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig2.pdf}
\caption{\textbf{Mean \textit{Grevillea australis} seedling density with burning, fire severity, adult density, altitude and Topographic Wetness Index (TWI) gradients.} Top left panel = centered and standardized model coefficients (on the log scale). Top right panel is the effect of burning. All other panels are partial dependence plots for each covariate along their observed gradients in areas burnt by the 2003 fires. All errors bars are 95\% Bayesian Credible Intervals.}
\end{minipage}
\end{figure}

\begin{figure}
\label{fig3}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig3.pdf}
\caption{\textbf{Observed and projected growth trajectories of four dominant Australian alpine shrubs.} Top panels: Thick lines represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. Bottom panels are the mean ($\pm$ 95\% Bayesian Credible Interval) projected height growth for each species over 70 years. In all cases, red or orange lines = seedlings growing in warmed (OTC) conditions and blue or light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. 34, 65, 65 \& 70 cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Prostanthera} \& \textit{Phebalium}, respectively).}
\end{minipage}
\end{figure}

\begin{figure}
\label{fig4}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig4.pdf}
\caption{\textbf{Mean ($\pm$ 95\% Bayesian Credible Interval) seedling probability of mortality for each species growing in control (blue) and warmed (red) conditions.}}
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\label{fig5}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig5.pdf}
\caption{\textbf{Experimental warming and inter-tussock gap size effects on \textit{Grevillea australis} seedlings.}First row = Growth rate effects, Second row = Mortality effects. Left panels are the centered and standardized model coefficients; right panels are the predicted growth and mortality rates along an inter-tussock gap size gradient warmed (red) and control (blue) conditions. All errors bars are 95\% Bayesian Credible Intervals.} 
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\label{fig6}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig6.pdf}
\caption{\textbf{Effects of warming on rates of tussock grass gap infilling.} Left panel = Model coefficients, Right panel = Predicted temporal change in size for an average 10 cm inter-tussock gap in warmed (red) and unwarmed (blue) conditions.} 
\end{minipage}
\end{figure}

>>>>

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment into arctic and alpine grasslands. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}) and that warming doubled tall shrub seedling growth rates and may increase survival. Warming had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increase survival.
\end{abstract}

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (two dominant shrubs of open heathland). In 2012-13, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a 200 m altitudinal range (equivalent to a 1.5\textdegree C change in mean temperature). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

Knowing shrub seedlings were more abundant in burnt vegetation, we investigated whether warmer post-fire conditions affected seedling growth and mortality rates. Seedlings of \textit{Grevillea}, \textit{Asterolasia}, \textit{Prostanthera} and \textit{Phebalium} emerging after a wildfire were transplanted into experimentally burnt plots that were either passively warmed by \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C using open top chambers (OTCs) or subject to current ambient conditions (controls). After \Sexpr{n_days(chamber_dates, FALSE)} days \Sexpr{n_days(chamber_dates, TRUE)} growing season days where OTC treatment was applied) or \Sexpr{round(n_days(chamber_dates, FALSE)/365.25)} years growth, mean seedling heights of the tall shrubs (\textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium}) growing in post-fire bare ground were taller in warmed plots by \Sexpr{ht_diff(non_tussock_otc_growth_data,'Grevillea')}, \Sexpr{ht_diff(non_tussock_otc_growth_data,'Prostanthera')} and \Sexpr{ht_diff(non_tussock_otc_growth_data,'Phebalium')} cm respectively (Fig. \ref{fig3}). Warming increased heights of both \textit{Grevillea} and \textit{Prostanthera} seedlings in all years, while \textit{Phebalium} did not respond to warming treatment until the second growing season. By contrast, seedlings of the prostrate shrub, \textit{Asterolasia}, showed no significant difference in growth rate between warmed and control plots in any year. For each species, we also observed similar treatment trends with stem diameter growth (Supplementary Fig. S3). Accounting for initial height and assuming logistic growth, mean annual predicted height growth rates of \textit{Grevillea}, \textit{Prostanthera} and \textit{Phebalium} were \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Grevillea')}, \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Prostanthera')} and \Sexpr{growth_rate_ratio(non_tussock_otc_growth_model,'Phebalium')} times that observed in control plots, respectively. Projecting these differences equates to an average 6 cm tall seedling (the mean initial size used in this experiment) growing in warmed conditions reaching maximum mature height in \Sexpr{hmax(non_tussock_otc_growth_model,'Grevillea')} fewer years for \textit{Grevillea} and \Sexpr{hmax(non_tussock_otc_growth_model,'Prostanthera')} fewer years for both \textit{Prostanthera} and \textit{Phebalium} (Fig. \ref{fig3}). \textit{Asterolasia} was predicted to reach a mature height in approximately \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[2])} to \Sexpr{round(hmax(non_tussock_otc_growth_model,'Asterolasia', FALSE)[1])} years regardless of warming treatment.

We validated the \textit{Grevillea} and \textit{Asterolasia} growth responses observed in experimentally warmed plots by using an elevation gradient in open heathland and maximum height data for 2003 post-fire recruits (i.e. suspected 9 year old seedlings). Here, a 200 m altitudinal range is equivalent to an ambient temperature difference of approximately \Sexpr{0.77 * (200/100)}\textdegree C\cite{Brown:1989vx}, which is comparable to that observed between experimental warmed and control plots (\Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C). In response to shifts in temperature, maximum heights of Grevillea post-fire recruits were expected to decrease with elevation, whereas \textit{Asterolasia} seedlings were not expected to show this pattern. Our experimental predictions were verified (Supplementary Fig. S4). Mean maximum heights of \textit{Grevillea} seedlings in burnt open heathland were \Sexpr{max_ht_alt(greaus_seedling_max_ht_model)} cm taller at 1670 m a.s.l compared to seedlings at 1860 m a.s.l. (\Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[1]} cm vs \Sexpr{max_ht_alt(greaus_seedling_max_ht_model, FALSE)[2]} cm; a difference comparable to our experimental findings). In contrast, mean maximum heights of \textit{Asterolasia} seedlings did not vary significantly with elevation. Topographic Wetness Index (TWI) and fire severity had no detectable influence on maximum seedling heights in either species.

@@

After five years and across all plots, \Sexpr{total_deaths(non_tussock_otc_mortality_data,TRUE)}\% (\Sexpr{total_deaths(non_tussock_otc_mortality_data,FALSE)} out of 511) of all seedlings growing in burnt bare ground patches had died. Most deaths occurred in \textit{Asterolasia} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Asterolasia')}) followed by \textit{Phebalium} (\Sexpr{death_counts(non_tussock_otc_mortality_data, 'Phebalium')}), \textit{Grevillea} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Grevillea')}) and \textit{Prostanthera} (\Sexpr{death_counts(non_tussock_otc_mortality_data,'Prostanthera')}). \textit{Prostanthera} showed the largest treatment effect (Fig. \ref{fig4}), with annual mortality rates estimated to be near \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','otc')}\% in warmed plots and \Sexpr{annual_mort(non_tussock_otc_mortality_model, 'Prostanthera','ctl')}\% in control plots. This significant increase in survival may be a consequence of OTCs reducing the severity of spring frosts by rising minimum ambient and soil temperatures by \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean_min','grand_mean')}\textdegree C and \Sexpr{micro_clim(overall_microclimate_summary,'CTemp_3cmBG','mean_min','grand_mean')}\textdegree C, respectively (see Supplementary Fig S5-7). Warming also reduced mean seedling mortality in \textit{Grevillea} and \textit{Phebalium} (Fig. \ref{fig4}); however, for both species, the effect was not statistically significant. By contrast, annual mortality rates in the prostrate shrub, \textit{Asterolasia}, were marginally higher in warmed plots, but again this effect was not significant (Fig. \ref{fig4}).

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still reduce shrub recruitment opportunities by infilling post-fire bare ground gaps at a faster rate. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses. However, the rate at which this occurred was not significantly influenced by a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C increase in temperature (Fig. \ref{fig6}).

@@

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig1.pdf}
\caption{\textbf{Conceptual diagram illustrating positive warming-shrub-fire feedback loop.} Top panel highlights the known positive feedback loop between fire, climatic warming and landscape flammability (solid blue lines). It also highlights several interactions (dashed \& black) which currently have a paucity of information but which may exacerbate or diminish this feedback loop. Bottom panel either assigns the direction or removes these interactions based on new evidence obtained from experimental and field surveys conducted in this study. It also includes an example of a potential, but unlikely, effect (red dotted line) that could break this feedback.}
\label{fig1}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig2.pdf}
\caption{\textbf{Mean \textit{Grevillea australis} seedling density with burning, fire severity, adult density, altitude and Topographic Wetness Index (TWI) gradients.} Top left panel = centered and standardized model coefficients (on the log scale). Top right panel is the effect of burning. All other panels are partial dependence plots for each covariate along their observed gradients in areas burnt by the 2003 fires. All errors bars are 95\% Bayesian Credible Intervals.}
\label{fig2}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig3.pdf}
\caption{\textbf{Observed and projected growth trajectories of four dominant Australian alpine shrubs.} Top panels: Thick lines represent mean ($\pm$ 95\% confidence intervals) observed heights at each May census. Thin lines represent individual growth trajectories. Bottom panels are the mean ($\pm$ 95\% Bayesian Credible Interval) projected height growth for each species over 70 years. In all cases, red or orange lines = seedlings growing in warmed (OTC) conditions and blue or light blue lines = seedlings growing in control conditions. Projections were based on an logistic growth model using initial seedling size of 6 cm (the mean initial height observed in the OTC experiment) and mean maximum heights observed in long-unburnt sites (i.e. 34, 65, 65 \& 70 cm for \textit{Asterolasia}, \textit{Grevillea}, \textit{Prostanthera} \& \textit{Phebalium}, respectively).}
\label{fig3}
\end{minipage}
\end{figure}

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig4.pdf}
\caption{\textbf{Mean ($\pm$ 95\% Bayesian Credible Interval) seedling probability of mortality for each species growing in control (blue) and warmed (red) conditions.}}
\label{fig4}
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig5.pdf}
\caption{\textbf{Experimental warming and inter-tussock gap size effects on \textit{Grevillea australis} seedlings.}First row = Growth rate effects, Second row = Mortality effects. Left panels are the centered and standardized model coefficients; right panels are the predicted growth and mortality rates along an inter-tussock gap size gradient warmed (red) and control (blue) conditions. All errors bars are 95\% Bayesian Credible Intervals.} 
\label{fig5}
\end{minipage}
\end{figure}
\pagebreak

\begin{figure}
\begin{minipage}{\textwidth}
\includegraphics{figures/fig6.pdf}
\caption{\textbf{Effects of warming on rates of tussock grass gap infilling.} Left panel = Model coefficients, Right panel = Predicted temporal change in size for an average 10 cm inter-tussock gap in warmed (red) and unwarmed (blue) conditions.} 
\label{fig6}
\end{minipage}
\end{figure}
